---
output:
  pdf_document: default
---

**Question 4** - Letting μt:y be the mean percentage cover for treatment level t in year y test the following hypothesis. Use a randomisation test in each case with 49 and then again with 499 resamples. [15 marks]


```{r, results = 'hide', eval=TRUE}
library(tidyverse)
library(dplyr)
library(scales)
library(readr)
```
```{r, echo=FALSE}
url <- "C:/Users/omcla/Documents/Test/biosci738_assignment_01/dicots_proportions.csv" #obtains data from Local
data <- read_csv(url) #places data into variable "data"
```
```{r, eval=FALSE}
## url <- "https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/dicots_proportions.csv" #obtains data from GitHub URL
## data <- read_csv(url) #places data into variable "data"
```

Code chunk for Data Manipulation

```{r, results = 'hide', eval=TRUE}
#extract the calluna data from the dataset
callunavulgaris_data <- data %>% select(starts_with("Calluna"), "Treat!")

#save dataframe in project folder
save(callunavulgaris_data, file = "callunavulgaris_data.RData")

# pivot long rather than wide 
calluna_long <- pivot_longer(callunavulgaris_data, 
                            cols = starts_with("Calluna"), 
                                     names_to = "Year", values_to = "Spread")

## rename columns to match the dataset 
names(calluna_long)[1] <- 'Treatment'

#find the index (place in the column) where the "Year" column ends with "08"
index_08 <- which(endsWith(calluna_long$Year,"08"))
print(index_08)

#for every index in the "Year" column. Change it to "2008"
for (index in index_08){
  calluna_long$Year[index] <- "2008"
}

#find the index (place in the column) where the "Year" column ends with "09"
index_09 <- which(endsWith(calluna_long$Year,"09"))
print(index_09)

#for every index in the "Year" column. Change it to "2009"
for (index in index_09){
  calluna_long$Year[index] <- "2009"
}

#find the index (place in the column) where the "Year" column ends with "10"
index_10 <- which(endsWith(calluna_long$Year,"10"))
print(index_10)

#for every index in the "Year" column. Change it to "2010"
for (index in index_10){
  calluna_long$Year[index] <- "2010"
}

#find the index (place in the column) where the "Year" column ends with "12"
index_12 <- which(endsWith(calluna_long$Year,"12"))
print(index_12)

#for every index in the "Year" column. Change it to "2012"
for (index in index_12){
  calluna_long$Year[index] <- "2012"
}

# wrangle the spread into percentage format
calluna_long <- calluna_long %>% 
  select(Treatment, Year, Spread)%>%
  mutate(     
    Spread = Spread * 100)

# find the mean and sd/se for the treatment type and year 
calluna_mean <- calluna_long %>% 
  group_by(Treatment, Year) %>% 
  summarise(., mean_year_treatment = mean(Spread), sd = sd(Spread), n = n(), se = sd / sqrt(n))
calluna_mean$mean_year_treatment = as.numeric(calluna_mean$mean_year_treatment)


## wrangle the data to have only 2008 and 2012 

calluna_8_12 <- calluna_long[!(calluna_long$Year == '2009' |calluna_long$Year == '2010'),]


## subset into B and C dataframes 

calluna_B <- subset(calluna_8_12, Treatment == 'B')
calluna_C <- subset(calluna_8_12, Treatment == 'C')
```

H0: mean % cover for Control treatment 2012 = mean % cover for Control treatment 2008
H1: mean % cover for Control treatment 2012 /=/ mean % cover for Control treatment 2008

```{r echo=FALSE}
## randomisation test for Control subset 49 

diff_in_means_C <- (calluna_C %>% group_by(Year)%>% 
  summarise(mean_C = mean(Spread)) %>% 
  summarise(diff = diff(mean_C)))$diff

diff_in_means_C
```

```{r, echo = FALSE}

nreps_49_C <- 49 # number of times I want to randomise

#empty array
randomarray_49_C <- numeric(nreps_49_C)
for (i in 1:nreps_49_C) {
  randomised_49C <- data.frame(value = calluna_C$Spread) ##the obersvations
  
  
  randomised_49C$random_labels <- sample(calluna_C$Year, replace = FALSE) ## randomise labels
  
  randomarray_49_C[i] <- as.list(randomised_49C %>% 
                                   group_by(random_labels)%>% summarise(mean = mean(value)) %>% ## randomised diff in mean
                                   summarise(diff = diff(mean)) )
}

## results
results_49_C <- data.frame(results = c(randomarray_49_C, diff_in_means_C))
results_49_C <- pivot_longer(results_49_C, cols = everything())

## calculate the p value 
n_exceed_49C <- sum(abs(results_49_C$value) >= abs(diff_in_means_C))
n_exceed_49C/nreps_49_C ## this is my p value for 49 reps of randomisation of B

```

```{r, echo = FALSE}
## randomisation test for Control subset 499 times 

nreps_499_C <- 499 # number of times I want to randomise

#empty array
randomarray_499_C <- numeric(nreps_499_C)
for (i in 1:nreps_499_C) {
  randomised_499C <- data.frame(value = calluna_C$Spread) ##the obersvations
  
  
  randomised_499C$random_labels <- sample(calluna_C$Year, replace = FALSE) ## randomise labels
  
  randomarray_499_C[i] <- as.list(randomised_499C %>% 
                                    group_by(random_labels)%>% summarise(mean = mean(value)) %>% ## randomised diff in mean
                                    summarise(diff = diff(mean)) )
}

## results
results_499_C <- data.frame(results = c(randomarray_49_C, diff_in_means_C))
results_499_C <- pivot_longer(results_499_C, cols = everything())
print(results_499_C$value)

## calculate the p value 
n_exceed_499C <- sum(abs(results_49_C$value) >= abs(diff_in_means_C))
n_exceed_499C/nreps_499_C ## this is my p value for 49 reps of randomisation of B
```



H0: mean % cover for Biocontrol treatment 2012 = mean % cover for Biocontrol treatment 2008
H1: mean % cover for Biocontrol treatment 2012 /=/ mean % cover for Biocontrol treatment 2008

```{r, echo=FALSE}
## randomisation test for BIOCONTROL subset 49 

diff_in_means_B <- (calluna_B %>% group_by(Year)%>% 
                      summarise(mean_B = mean(Spread)) %>% 
                      summarise(diff = diff(mean_B)))$diff

diff_in_means_B
```

```{r, echo=FALSE}
nreps_49_B <- 49 # number of times I want to randomise

#empty array
randomarray_49_B <- numeric(nreps_49_B)
for (i in 1:nreps_49_B) {
  randomised_49B <- data.frame(value = calluna_B$Spread) ##the obersvations
  
  
  randomised_49B$random_labels <- sample(calluna_B$Year, replace = FALSE) ## randomise labels
  
  randomarray_49_B[i] <- as.list(randomised_49B %>% 
                                   group_by(random_labels)%>% summarise(mean = mean(value)) %>% ## randomised diff in mean
                                   summarise(diff = diff(mean)) )
}

## results
results_49_B <- data.frame(results = c(randomarray_49_B, diff_in_means_B))
results_49_B <- pivot_longer(results_49_B, cols = everything())

## calculate the p value 
n_exceed_49B <- sum(abs(results_49_B$value) >= abs(diff_in_means_B))
n_exceed_49B/nreps_49_B ## this is my p value for 49 reps of randomisation of B

```

```{r, echo=FALSE}
## randomisation test for BIOCONTROL subset 499 

nreps_499_B <- 499 # number of times I want to randomise

#empty array
randomarray_499_B <- numeric(nreps_49_B)
for (i in 1:nreps_499_B) {
  randomised_499B <- data.frame(value = calluna_B$Spread) ##the obersvations
  
  
  randomised_499B$random_labels <- sample(calluna_B$Year, replace = FALSE) ## randomise labels
  
  randomarray_499_B[i] <- as.list(randomised_499B %>% 
                                    group_by(random_labels)%>% summarise(mean = mean(value)) %>% ## randomised diff in mean
                                    summarise(diff = diff(mean)) )
}

## results
results_499_B <- data.frame(results = c(randomarray_499_B, diff_in_means_B))
results_499_B <- pivot_longer(results_499_B, cols = everything())

## calculate the p value 
n_exceed_499B <- sum(abs(results_49_B$value) >= abs(diff_in_means_B))
n_exceed_499B/nreps_499_B ## this is my p value for 49 reps of randomisation of B
```

Observed Test Statistic for Permutation Tests:

Control: 
```{r echo=FALSE}
print(diff_in_means_C)
```

Biocontrol:
```{r, echo=FALSE}
print(diff_in_means_B)
```

Permutation Tests

P-value for Control:
```{r, echo=FALSE}
## PERMUTATION TESTS

#control 
combinations_C <- combn(12,6)
permtest_combn_C <- apply(combinations_C, 2, function(x)
  mean(calluna_C$Spread[x]) - mean(calluna_C$Spread[-x]))
p_val_C <- length(permtest_combn_C[abs(permtest_combn_C) >= diff_in_means_C]) / choose(12,6)
p_val_C
```

P-value for Biocontrol:
```{r, echo=FALSE}
#biocontrol
combinations_B <- combn(12,6)
permtest_combn_B <- apply(combinations_B, 2, function(x)
  mean(calluna_B$Spread[x]) - mean(calluna_B$Spread[-x]))
p_val_B <- length(permtest_combn_B[abs(permtest_combn_B) >= abs(diff_in_means_B)]) / choose(12,6)
p_val_B
```

The above p-values show what you would expect from the results - that the control does not show a low p-value, i.e. the result we received was statistically likely to occur. Therefore we accept the null hypothesis that there is no difference in means between the controls. 
Conversely, the p-value for the biocontrol subset supports the hypothesis that there is a difference between the means between 2008 and 2012, showing the biocontrol treatment had a significant effect on the percentage of ground cover of heather. 


**Histograms**

Control:

```{r, echo=FALSE}


## HISTOGRAM

hist(permtest_combn_C,
  xlab='Differences Between Randomised Group Means',
  main='Permutation Test (Control)')
abline(v = diff_in_means_C, col = "red")


hist(permtest_combn_B,
     xlab='Differences Between Randomised Group Means',
     main='Permutation Test (Biocontrol)')
abline(v = diff_in_means_B, col = "red")
```


From the histograms I think it backs up the conclusions drawn from the p-values. The control histogram clearly shows the observed test statistic within the normal range of the expected observations, ie an expected result within the null hypothesis. 

The histogram for the biocontrol subset, however, shows the observed test statistic outside of the normal range of expected results showing that something has acted to disprove the null hypothesis (ie the biocontrol treatment). 